Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2018-08-20 documentation/paper.tex
Renamed Acoustic Parameter to AP

The Acoustic Parameter method (abbreviated AP) is the technique that was used for event discrimination in the original PICO-60 analysis. A function was defined that discriminates between alpha particles and nuclear recoils based on audio data. It was derived based on a physical model of the bubble chamber, and was tuned using neutron calibration sources to produce events in the PICO-60 apparatus.

Before the Acoustic Parameter can be calculated, the audio data must first be converted into the frequency domain (specifically, with the banded Fourier transform $\beta_{8}$). It must also be preprocessed with the position correction function $PosCor(\beta_{8}, \chi)$, which adjusts the overall volume of an audio recording according to its position in the vessel. This is done because the acoustic characteristics differ significantly depending on the position of an event in the vessel.

Finally, a set of data cuts must be applied to remove certain classes of events which are not accurately classified by Acoustic Parameter. These classes include wall event and multiple-bubble events.

@@

Beyond this, the banded Fourier transform $\beta_{8}$ (which is the input to Acoustic Parameter and the original neural analysis) was computed by integrating the resonant energy over all frequencies within each of a set of bands. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

@@

\begin{itemize}
    \item The run was not collected during engineering or testing
    \item Recording was triggered by the camera
    \item Acoustic Parameter is not erroneously large and negative ($log_{10}(AP)>-100$)
    \item Recorded more than 25 seconds after reaching target pressure
    \item The bubble position $\chi$ was successfully calculated ($[\chi_{X}, \chi_{Y}, \chi_{Z}]\neq[-100, -100, -100]$)
\end{itemize}

@@

Events that occur near the walls of the vessel have acoustic properties which are notably different from events nearer the center of the vessel. It can be desirable for a discriminator to handle these events correctly; however, Acoustic Parameter does not, and neither does the neural network used in the previous PICO-60 paper. Thus, removing wall events allowed for a more meaningful direct comparison between a new neural network and existing techniques.

@@

One of the key benefits of machine learning systems is that they have the capacity to approximate arbitrary functions (such as Acoustic Parameter) without human intervention. This greatly alleviates the need for human programmers to define and optimize the exact algorithm used.

@@

The first step is to calculate the standard deviation S of the union of the two sets. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation, Acoustic Parameter has a significantly wider range. Normalization of the union prevents this from creating a bias where Acoustic Parameter would produce a higher standard deviation with a similarly proportioned error.

@@

All supervised learning configurations were trained and evaluated based on their ability to determine the origin of events: whether they are from calibration runs (predominantly neutrons) or background runs (predominantly alpha particles). This metric was used for training and testing instead of Acoustic Parameter for two reasons:

\begin{enumerate}
    \item In future experiments, when no Acoustic Parameter is available, impure data is the only information available for training. The network structure should be selected to accommodate this.
    \item In general, supervised learning systems learn to replicate biases present in the training data set. Thus, to evaluate the network's effectiveness at processing the input data, the run type (which is used for training) must be used as a baseline instead of Acoustic Parameter.
\end{enumerate}

@@

$WallCut(D)$ was applied to the low-resolution Fourier transforms during early experiments (since they were used for Acoustic Parameter) to minimize the number of variables between initial neural networks and a proven discriminator.

@@

\begin{enumerate}
    \item This was a highly effective configuration with wall cuts, producing 97\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as Acoustic Parameter makes the algorithms more directly comparable.
\end{enumerate}

@@

Semi-supervised learning is an uncommon machine learning technique, relative to widely-used supervised and unsupervised learning. The concept is to train a machine learning model on a set of labeled (classified) data in addition to a set of unlabeled (unclasified) data.

@@

Since the intent of these techniques is to separate particle types (the same as Acoustic Parameter), performance evaluations (accuracy and class-wise standard deviation) use Acoustic Parameter as a baseline.

@@

During a grid search of parameters of the gravitational differentiation algorithm (in addition to the stochastic gradient descent learning rate), the algorithm achieved a maximum of 98\% validation accuracy. The mean over that run (also the highest overall) was 96\%. Figure \ref{grav_grid_search} shows the highly accurate replication of Acoustic Parameter.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{grav_grid_search}
    \caption{\label{grav_grid_search} Accuracy of best gravitational differentiation training run compared to Acoustic Parameter (98\% accuracy).}
\end{figure}

@@

During a grid search of these parameters (in addition to regularization parameters of the network), the best accuracy achieved for replication of Acoustic Parameter was exactly 100\%. The mean of all epochs within that training run was 98\%.

As seen in Figure \ref{icn_grid_search}, the network learned to replicate Acoustic Parameter very accurately. Interestingly, the examples with Acoustic Parameter values between the neutron and alpha particle ranges were also classified as such by the network. This may imply the indecisiveness on these examples is fundamental to the data rather than an implementation problem with Acoustic Parameter.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{icn_grid_search}
    \caption{\label{icn_grid_search} Accuracy of optimal iterative cluster nucleation training run compared to Acoustic Parameter (100\% accuracy).}
\end{figure}

@@

The following table directly compares performance of the semi-supervised learning systems, with the existing PICO-60 neural network provided as a baseline once again. Note that ``Max AP Accuracy'' refers to accuracy at replication of Acoustic Parameter, rather than replication of the ground truths.

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like Acoustic Parameter. This conclusion has been achieved through two rounds of optimization, the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems and replicate Acoustic Parameter as accurately as possible.

@@

Two semi-supervised learning algorithms were developed, applying the best configuration found for supervised learning. While both techniques produced some very effective discriminators, iterative cluster nucleation was found to be the more effective of the two. It replicated Acoustic Parameter with a mean of 98\% accuracy on the best configuration. This indicates that the technique shows significant promise, and may be worth consideration as a discrimination technique for future experiments.

>>>>

The Acoustic Parameter method (furthermore AP) is the technique that was used for event discrimination in the original PICO-60 analysis. A function was defined that discriminates between alpha particles and nuclear recoils based on audio data. It was derived based on a physical model of the bubble chamber, and was tuned using neutron calibration sources to produce events in the PICO-60 apparatus.

Before AP can be calculated, the audio data must first be converted into the frequency domain (specifically, with the banded Fourier transform $\beta_{8}$). It must also be preprocessed with the position correction function $PosCor(\beta_{8}, \chi)$, which adjusts the overall volume of an audio recording according to its position in the vessel. This is done because the acoustic characteristics differ significantly depending on the position of an event in the vessel.

Finally, a set of data cuts must be applied to remove certain classes of events which are not accurately classified using AP. These classes include wall event and multiple-bubble events.

@@

Beyond this, the banded Fourier transform $\beta_{8}$ (which is the input to the AP function and the original neural analysis) was computed by integrating the resonant energy over all frequencies within each of a set of bands. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

@@

\begin{itemize}
    \item The run was not collected during engineering or testing
    \item Recording was triggered by the camera
    \item AP is not erroneously large and negative ($log_{10}(AP)>-100$)
    \item Recorded more than 25 seconds after reaching target pressure
    \item The bubble position $\chi$ was successfully calculated ($[\chi_{X}, \chi_{Y}, \chi_{Z}]\neq[-100, -100, -100]$)
\end{itemize}

@@

Events that occur near the walls of the vessel have acoustic properties which are notably different from events nearer the center of the vessel. It can be desirable for a discriminator to handle these events correctly; however, AP does not, and neither does the neural network used in the previous PICO-60 paper. Thus, removing wall events allowed for a more meaningful direct comparison between a new neural network and existing techniques.

@@

One of the key benefits of machine learning systems is that they have the capacity to approximate arbitrary functions (such as AP) without human intervention. This greatly alleviates the need for human programmers to define and optimize the exact algorithm used.

@@

The first step is to calculate the standard deviation S of the union of the two sets. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

@@

All supervised learning configurations were trained and evaluated based on their ability to determine the origin of events: whether they are from calibration runs (predominantly neutrons) or background runs (predominantly alpha particles). This metric was used for training and testing instead of AP for two reasons:

\begin{enumerate}
    \item In future experiments, when no AP equivalent is available, impure data is the only information available for training. The network structure should be selected to accommodate this.
    \item In general, supervised learning systems learn to replicate biases present in the training data set. Thus, to evaluate the network's effectiveness at processing the input data, the run type (which is used for training) must be used as a baseline instead of AP.
\end{enumerate}

@@

$WallCut(D)$ was applied to the low-resolution Fourier transforms during early experiments (since they were used for AP) to minimize the number of variables between initial neural networks and a proven discriminator.

@@

\begin{enumerate}
    \item This was a highly effective configuration with wall cuts, producing 97\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as AP makes the algorithms more directly comparable.
\end{enumerate}

@@

Semi-supervised learning is an uncommon machine learning technique, relative to widely-used supervised and unsupervised learning. The concept is to train a machine learning model on a set of labeled (classified) data in addition to a set of unlabeled (unclassified) data.

@@

Since the intent of these techniques is to separate particle types (the same as AP), performance evaluations (accuracy and class-wise standard deviation) use AP as a baseline. Note that $WallCut(D)$ was once again applied for semi-supervised learning, because $FourierMLP(PosCor(\beta_{8}))$ was used for a neural network configuration, and because AP requires application of $WallCut(D)$ to be meaningfully comparable to any neural network.

@@

During a grid search of parameters of the gravitational differentiation algorithm (in addition to the stochastic gradient descent learning rate), the algorithm achieved a maximum of 98\% validation accuracy. The mean over that run (also the highest overall) was 96\%. Figure \ref{grav_grid_search} shows the highly accurate replication of AP.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{grav_grid_search}
    \caption{\label{grav_grid_search} Accuracy of best gravitational differentiation training run compared to AP (98\% accuracy).}
\end{figure}

@@

During a grid search of these parameters (in addition to regularization parameters of the network), the best accuracy achieved for replication of AP was exactly 100\%. The mean of all epochs within that training run was 98\%.

As seen in Figure \ref{icn_grid_search}, the network learned to replicate AP very accurately. Interestingly, the examples with AP values between the neutron and alpha particle ranges were also classified as such by the network. This may imply the indecisiveness on these examples is fundamental to the data rather than an implementation problem with AP.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{icn_grid_search}
    \caption{\label{icn_grid_search} Accuracy of optimal iterative cluster nucleation training run compared to AP (100\% accuracy).}
\end{figure}

@@

The following table directly compares performance of the semi-supervised learning systems, with the existing PICO-60 neural network provided as a baseline once again. Note that ``Max AP Accuracy'' refers to accuracy at replication of AP, rather than replication of the ground truths.

@@

This study has demonstrated that it is possible, using semi-supervised learning, to automatically optimize a discriminator with performance comparable to one developed and tuned by humans like the Acoustic Parameter technique. This conclusion has been achieved through two rounds of optimization, the first to explore and evaluate various network structures and input formats using supervised learning, and the second to incorporate the most successful configuration into semi-supervised learning systems and replicate AP as accurately as possible.

@@

Two semi-supervised learning algorithms were developed, applying the best configuration found for supervised learning. While both techniques produced some very effective discriminators, iterative cluster nucleation was found to be the more effective of the two. It replicated AP with a mean of 98\% accuracy on the best configuration. This indicates that the technique shows significant promise, and may be worth consideration as a discrimination technique for future experiments.
